Northern Pikeminnow Predation on Juvenile Chinook Salmon in Deadwater Slough, Salmon River, Idaho and Potential Impacts to Local Populations

### 2022-01-10 analysis/paper/paper.Rmd
Added a few sentences about the degree of bias that could be introduced by difference in catchability between marked and unmarked fish

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling, and possibly biasing them low. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> We further recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would result in marked fish being less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability.

>>>>

All mark-recapture abundance estimators suggest a large number of Northern Pikeminnow occupying Deadwater Slough. Several assumptions in our estimators may have influenced the magnitude of our results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the limited length of the sampling period (two weeks), that assumption is likely met. The assumption of a closed population can also affect results if mortality occurs for some marked fish released back to the population. During our field processing, we did not observe mortalities or injury post release; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that the second (and subsequent) sample(s) are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size threshold susceptible to angling, and possibly biasing them low. We also consider the validity of our assumption of equal catchability of individual fish between sampling events to have introduced potential bias in our results. Unequal catchability for individual fish is difficult to assess, especially if being caught once influences the likelihood of recapture. During our study, we recaptured two individual fish during multiple surveys indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.-->

We further recognize that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit the "good" fishing locales. If marked fish were not recently released or mixed back into those areas that would result in marked fish being less susceptible to capture. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish, so if marked fish are only 70% as likely to be caught as unmarked fish (due to behavior or spatial distribution) the true abundance is closer to 70% of our abundance estimate. We calculated this for the Schnabel estimator by assuming that only 70% of the fish marked each sample were returned to the populuation and available for recapture.
